---
title: "Biostat 203B Homework 1"
author: Jiahao Tian
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=T}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email. You'll get GitHub Pro account for free (unlimited public and private repositories).

2. Create a **private** repository `biostat-203b-2022-winter` and add `Hua-Zhou` and `maschepps` as your collaborators with write permission.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `main` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `main` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and extra data sets to reproduce results) in `main` branch.

4. After each homework due date, teaching assistant and instructor will check out your main branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

## Q2. Data ethics training

This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic.mit.edu/docs/gettingstarted/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Solution:**  CITI URL

https://www.citiprogram.org/verify/?k5335f2a8-16bc-4351-bdc1-567d7d1de024-46572249

https://www.citiprogram.org/verify/?wf717362f-6e7e-426c-942e-465fa9c6ab46-46572249



## Q3. Linux Shell Commands

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from MIMIC-IV. Refer to the documentation <https://mimic.mit.edu/docs/iv/> for details of data files.  
    ```{bash}
    ls -l /mnt/mimiciv/1.0
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/mnt/mimiciv/1.0` directly in following exercises. 

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data files distributed as `.csv.gz` files instead of `.csv` (comma separated values) files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what's in each folder.

**Solution:** Content of the folder `core` is 
```{bash}
ls -l /mnt/mimiciv/1.0/core
```

**Solution:** Content of the folder `hosp` is
```{bash}
ls -l /mnt/mimiciv/1.0/hosp
```

**Solution:** Content of the folder `icu` is
```{bash}
ls -l /mnt/mimiciv/1.0/icu
```

**Solution:**
The reason for using `.cvs.gz` files is that these three files are large files, so when `.cvs.gz` can help to reduce the file sizes, it's worth it. Furthermore, `.cvs.gz` is easier to manage and store.

3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.

**Solution:**

a. `zcat` is a command allows the user to expand and view a compressed file without uncompressing that file.

b. `zless` is a command (filter) which allows examination of compressed or plain text files on screenful at a time on a soft-copy terminal. Can scroll upwards and downwards through the input.

c. `zmore` is a command (filter) which allows examination of compressed or plain text files on screenful at a time on a soft-copy terminal. Can only scroll downwards through the input.

d. `zgrep` is a command that is used to search the contents of a compressed file without uncompressing it.


4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
    ```

**Solution:** The output will list detailed contents of the `.gz` file under `/mnt/mimiciv/1.0/core/` directory.

   Display the number of lines in each data file using a similar loop.

**Solution:**

``` {bash}
   for datafile in /mnt/mimiciv/1.0/core/*.gz
   do
       zcat $datafile | wc -l
   done
```
   
  
5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, `uniq`, `wc`, and so on.)

**Solution:**

a. display first five lines

```{bash}
   zcat /mnt/mimiciv/1.0/core/*.gz | awk '(NR <= 5)'
```

**Solution:**

b. number of rows in this datafile

```{bash}
   zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk 'END { print NR }'
```

**Solution:**

c. number of unique patients

```{bash}
   zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | awk -F, '{ print $1 }' | sort | uniq | wc -l
```


6. What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)
cle
**Solution:** 
possible values of each variable
`admission_type`
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz |sed '1d'|
awk -F, '{print $6}'| sort | uniq -c
```


## Q4. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    ```

Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in Git. 

**Solution:**
`wget -nc` is the non-interactive network downloader which is used to download files from the server even when the user has not logged on to the system and it can work in the background without hindering the current process.


Complete the following loop to tabulate the number of times each of the four characters is mentioned using Linux commands.

**Solution:**

```{bash, eval=FALSE}
wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    for char in Elizabeth Jane Lydia Darcy
    do
     echo $char:
     gerp -o -i $char pg4267.txt | wc -l
    done
    
```



2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

**Solution**

Both command will create these two words "hello, world" to a file. For the `>` command might clean whatever in the file, but for `>>` just add these two word to a file without doing anything.


3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pg42671.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

### Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution:**

`cal` only shows the current year, current month, and today's day.

`cal 2021` shows the whole 12 months of 2021.

`cal 9 1752` shows the year 1752, the month of September.
